Abstract Control in Static Analysis

### 2014-09-21 writeup/darais-quals.tex
things are getting polished.  next to polish is proofs.

We use traditional notation for function definitions and applications rather than lambda calculus notation.
For example, we will write $f(w, x) ≔ g(y, z)$ rather than $f\; w\; x ≔ g\; y\; z$.

We use lambda notation for anonymous functions.
For example, defining the identify function could be notated $id ≔ λ(x) → x$ rather than the named definition $id(x) ≔ x$.

@@

A basic familiarity with the concept of monads will be useful for understanding the rest of the paper.
Some find it useful to review the concept of functor before learning about monads. 
We adopt this approach for our brief explanation of monads.

A type $𝒯 : Set → Set$ is called a \emph{functor} if one can:
\begin{itemize}
\item Define $map : ∀ (A : Set), (A → B) → (𝒯(A) → 𝒯(B))$
\item Prove $map(id) = id$, where $id$ is the identity function $λ(x) → x$
\item Prove $map (g ∘ f) = map(g) ∘ map(f)$
\end{itemize}

\paragraph{Example:} 
Lists are functors, where $map$ is defined:
\begin{align*}
    map(f)(xs) ≔ &\case(xs):           \\
          Nil    &→ Nil                \\
          x ∷ xs &→ f(x) ∷ map(f)(xs') \\
\end{align*}

@@

Likewise a type $𝒯 : Set → Set$ is called a \emph{monad} if one can:
\begin{itemize}
\item Define $extend : ∀ (A : Set), (A → 𝒯(B)) → (𝒯(A) → 𝒯(B))$
\item Prove unit and associativity laws (not mentioned here).
\end{itemize}
The only difference from the definition of functor is in the the first argument $(A → 𝒯(B))$.
For monads, this argument is allowed to be a \emph{monadic} function, rather than a pure function $(A → B)$.

\paragraph{Example:} Lists are monads, where $extend$ is defined:
\begin{align*}
    extend(f)(xs) ≔ &\case(xs):              \\
                Nil &→ Nil                   \\
            x ∷ xs' &→ f(x) ⧺ extend(f)(xs') \\
\end{align*}
$extend$ for lists can be understood as $map$ followed by the flattening operation $concat$.

@@

\paragraph{Example:} 
$extend$ for lists can also be understood as $concat$ followed by $map$:
\begin{align*}
extend(f)([10, 100]) &= concat(map(f)([10, 100]))    \\
                     &= concat([[9, 11], [99, 101]]) \\
                     &= [9, 11, 99, 101]             \\
\end{align*}
For all monads, it is equivalent to define $return$, $map$ and $join$, where $join : ∀ (A : Set), 𝒯(𝒯(A)) → 𝒯(A)$.
As we have just seen, $join$ for the list monad is just $concat$.

@@

Monad transformers are functions between monads.
Where a monad $𝒯$ will have type $Set → Set$, a monad transformer will have type $(Set → Set) → (Set → Set)$.
Perhaps a more appropriate type is $Monad → Monad$.

Monad transformers are used to add an effect on top of another monad.
The three monads used in this work are the state monad, powerset monad, and the identity monad.
The state monad is notated $\SM(𝓈)$ (carrying a single cell of type $𝓈$), the powerset monad is notated $\PM$, and the identity is notated $ID$.
The state and powerset monad transformer equivalents are notated $\ST(𝓈)(𝓂 )$ and $\PT(𝓂 )$ where $𝓂 $ is the underlying monad.
The state monad has $get$ and $put$ effects, whereas the powerset monad has a $nondeterminism$ effect.
The transformer versions of both of these monads allow you to combine effects piecewise.

\paragraph{Example:}
$\ST(ℤ)(\PT(ID))$ is a monad which has $get$ and $put$ effects for a single integer state, in addition to nondeterminism effects.

@@

This work follows the small-step operational approach to language semantics pioneered independently by Plotkin and Felleisen.
In small-step operational semantics, the meaning of an expression $e$ in a language is given by the transitive closure of some step relation $e ↦ e'$.
This yields a state-machine approach to semantics.
Exploring the meaning of $e$ reduces to exploring all possible transitions reachable from $e$ through $↦$.
This contrasts to previously developed deontational approaches, where the meaning of $e$ would be given by a denotation function $⟦ e ⟧$.
This denotation function can quickly demand lots of mathematical machinery, and the small-step operational method side-steps this issue.

@@

Abstract Iterpretation (AI) is a formal framework for program analysis pioneered by Cousot and Cousot.
In the setting of abstract interpretation, a program analysis is just an alternate semantics over an abstract domain.
To distinguish the original semantics and the alternate semantics, we call the original semantics the \emph{concrete} semantics, and the alternate semantics the \emph{abstract} semantics\footnote{
  This terminology quickly becomes confusing because we will eventually construct an abstract machine for both semantics, yielding a concrete abstract machine and an abstract abstract machine. 
  This pun is exploited mercilessly in the work of Might and Van Horn as they "abstract" abstract abstract machines, yielding an abstract abstract abstract machine.
}.

Consider a concrete language $\C$ and a program in that language $(\Ce : \C)$.
In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A relationship between $\C$ and $\A$ that defines when $(\Ce : \C)$ and $(\Ae : \A)$ are related.
  This often takes the form of projecting $\A$ to $\PM(\C)$ and using the subset relation.
\item 
  An abstract version of $\Ce$ that lives in $\A$: $(\Ae : \A)$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A way to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
By convention, we put hats ($\widehat{}$) on things to name their abstract cousins.
The content of the approach is shown in the following picture:
\input{diagrams/AI.tikz}
This picture takes place at some point in the analysis process.
$\Ce$ is the current program (possibly the result of running the base program for a little while).
$\Ae$ is a valid abstraction of $\Ce$, where this validity is expressed some relation $R$ holding between $\Ce$ and $\Ae$.
$\Ce'$ is some next state of execution of $\Ce$, and likewise for $\Ae$/$\Ae'$.
In order to be a correct analysis, it must be guaranteed that $\Ce'$ and $\Ae'$ will be related.
The logical structure of the picture is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce R \Ae) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' R \Ae')
\end{equation*}

@@

The AI setting can be elegently simplified and enriched through the use of \emph{galois connections}.
Galois connections serve as a unifying framework for establishing the "relationship between $\C $ and $\A$" mentioned in the previous section.

A galois connection between two posets (sets with a partial order) $\C$ and $\A$ is notated $\C\galois{α}{γ}\A$ and contains:
\begin{itemize}
\item $(α : \C → \A)$ where $α$ is monotonic
\item $(γ : \A → \C )$ where $γ$ is monotonic
\item $(γ ∘ α)$ is expansive: $∀ (x : \C ), x ⊑ γ(α(x))$
\item $(α ∘ γ)$ is contractive: $∀ (y : \A), α(γ(y)) ⊑ y$
\end{itemize}
The last two properties can be succinctly stated as $(α ∘ γ ⊑ id ⊑ γ ∘ α)$\footnote{
  This uses the logical monotonicity relation $f ⊑ g ⇔  (x ⊑ y ⇒  f(x) ⊑ g(y))$ for the function space
}.
Equivalent to all four properties is the property $x ⊑ γ(y) ⇔  α(x) ⊑ y$.

\paragraph{Example:}
Given a galois connection $\C\galois{α}{γ}\A$, there exists a galois connection $\C → \C  \galois{α}{γ} \A → \A$ where:
\begin{align*}
α(f : \C → \C) &≔ γ ∘ f ∘ α \\
γ(g : \A → \A) &≔ α ∘ g ∘ γ \\
\end{align*}

\paragraph{Example:} 
The language $(\PM(ℤ),+,*)$ forms a galois connection with the language $(\PM({EVEN, ODD}),∧,∨)$ where:
\begin{align*}
α(zs : \PM(ℤ))           &≔ ⋃ { { EVEN | ∃ z ∈ zs ∧ Even(z)}, { ODD | ∃ z ∈ zs ∧ Odd(z) } }      \\
γ(ts : \PM({EVEN, ODD})) &≔ ⋃ { { z ∈ ℤ | EVEN ∈ ts ∧ Even(z) }, { z ∈ ℤ | ODD ∈ ts ∧ Odd(z) } } \\
\end{align*}

Galois connections simplify the AI framework by using $x ⊑ γ(y)$ or (equivalently) $α(x) ⊑ y$ as the relation $(x R y)$.
Put differently, galois connections are a natural and general way of placing partial orders \emph{on sets themselves} (note that x and y live in different sets).
One can also think of galois connections as something like an isomorphism, but with a weaker round-trip property.
(An isomorphism would require $α ∘ γ = id = γ ∘ α$.)

Using galois connections, the AI framework introduced in the previous section can be re-stated. In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A galois connection $\C\galois{α}{γ}\A$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A way to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
The content of the approach is shown in the following picture:
\input{diagrams/AIGC.tikz}
This picture takes place at some point in the analysis process.
$\Ce$ is the current program (possibly the result of running the base program for a little while).
$\Ae$ is a valid abstraction of $\Ce$ by construction as $α(\Ce)$.
$\Ce'$ is some next state of execution of $\Ce$, and likewise for $\Ae$/$\Ae'$.
In order to be a correct analysis, it must be guaranteed that $\Ce' ⊑ γ(\Ae')$.
The logical structure of the picture is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce ⊑ γ(\Ae)) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' ⊑ γ(\Ae'))
\end{equation*}

The logical structure of the previous figure is the \emph{soundness} property of the analysis.
This property can be simplified further by the statement $\Csteps ⊑ γ(\Asteps)$.
This uses the definition of galois connections for function spaces shown in a previous example.
What is more, this more compact statement of soundness also captures another property, the \emph{tightness} of the analysis.
Unfolded, this property states:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce' ⊑ γ(\Ae')) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce ⊑ γ(\Ae))
\end{equation*}
Tightness tells us that our abstraction is provably \emph{the best abstraction possible}.
The beauty of galois connections is how they capture both soundness and tightness of an analysis with one concise statement: $\Csteps ⊑ γ(\Asteps)$.

@@

We define the CPS-IF language as follows:
\begin{align*}
x,y,k &: Var  &⩴ ..variables..                          \\
b,i,l &: Lit  &⩴ ℤ ⋃ 𝔹                                  \\
f,g   &: Lam  &⩴ λ(x) → c | λ(x,k) → c                  \\
a     &: Atom &⩴ x | l | f | op(a)                      \\
op    &: Op   &⩴ add1 | sub1 | gez                      \\
c     &: Call &⩴ if(a){c}{c} | a(a) | a(a,a) | halt(a)  \\
\end{align*}

@@

Control flow analysis is a class of analysis which is particularly important for higher-order languages.
In non-higher-order languages, it is useful to distinguish control-flow--which functions can be called--from data-flow--which values flow where.
Traditionally one performs a control-flow analysis first, to find out which functions are called, and a data-flow analysis second, using the results of the control flow analysis.

In higher-order languages, data-flow and control-flow become the same problem.
Before you can tell which functions will be called, you need to know how values flow, because functions are themselves values.

We note that the distinction between "higher-order" languages and "non-higher-order" languages is a red herring.
Functions can be passed as values in C using function pointers, and object-oriented languages enjoy a similar circularity between control- and data- flow from method dispatch.
You can read these statements as saying "all languages are higher order, and thus need control flow analysis".
Or you can read them as saying "all languages are higher order, and we've done just fine without control flow analysis so far".
The distinction is in the code that you end up analysing.

We build on the tradition of control flow analysis pioneered by Shivers and its recent refinements developed by Might and Van Horn.
In this tradition, the result of a control flow analysis is closer to what one would think of as a data-flow analysis.
0CFA, the most basic control flow analysis, computes (conservatively) the set of values which might flow to a particular variable.
kCFA is a suite of context sensitive extensions to 0CFA which analyses a given function separately for each possible calling context.
We use 0CFA as the example analysis in this paper.
However our framework, and our implementation of it, scale to the full range of context sensitive control flow analyses.

@@

Monadic AI (Sergey et al. PLDI 2013) showed that computational modularity for abstract interpreters can be achieved through a monadic abstraction.
We build directly on this work at the level of intuition--monads serve as our pivot-point for generalizing abstract control.
However, in our pursuit of an axis-independent correctness framework, our approach deviates greatly from this work in the specific interfaces used.

First we demonstrate a simple 0CFA analysis on CPS-IF which doesn't use our framework.
The language is the same as introduced in section \ref{section:Background:ContinuationPassingStyle}, and repeated here for convenience:
\begin{align*}
x,y,k &: Var  &≔ ..variables..                          \\
b,i,l &: Lit  &≔ ℤ ⋃ 𝔹                                  \\
f,g   &: Lam  &⩴ λ(x) → c | λ(x,k) → c                  \\
a     &: Atom &⩴ x | l | f | op(a)                      \\
op    &: Op   &⩴ add1 | sub1 | gez                      \\
c     &: Call &⩴ if(a){c}{c} | a(a) | a(a,a) | halt(a)  \\
\end{align*}
The abstract semantics of 0CFA will track literals (including lambdas) that appear in the program text.
For integers that do not appear in the program text, the analysis uses a single token $INT$ which conservatively approximates any possible integer.
Because there are a finite many literals in the program text, there are a finite number of possible values.
Formally, the state space for the abstract machine of the abstract semantics is defined as:
\begin{align*}
v &: \AVal   ⩴ INT | l | λ(x) → c | λ(x,k) → c \\
σ &: \AStore ≔ Var → \PM(\AVal)                 \\
Σ &≔ \PM(Call × \AStore)                        \\
\end{align*}

The abstract semantics for $Σ$ comes in two parts.  
First we define a \emph{denotation function} $(𝒜 : \AStore × Atom → \PM(\AVal))$ for $Atom$ expressions.
Second we define a \emph{step relation} (as a function) $(𝒞 : Call × \AStore → \PM(Call × \AStore))$ for $Call$ expressions.
\begin{align*}
𝒜                            &: \AStore × Atom → \PM(\AVal)  \\
𝒜 (σ,x)                      &≔ σ(x)                        \\
𝒜 (σ,l)                      &≔ { l }                       \\
𝒜 (σ,add1 a) | 𝒜 [ρ,σ] ⊑ INT &≔ { INT }                     \\
𝒜 (σ,sub1 a) | 𝒜 [ρ,σ] ⊑ INT &≔ { INT }                     \\
𝒜 (σ,gez a)  | 𝒜 [ρ,σ] ⊑ INT &≔ { T , F }                   \\
𝒜 (σ,λ(x)(k) → c)            &≔ { λ(x)(k) → c }             \\
\end{align*}

% \begin{align*}
% 𝒞                   &: &Call × \AStore → \PM(Call × \AStore)                                     \\
% 𝒞 <if(a){c₁}{c₂},σ> &≔ &{ <c,σ>                                                                 \\
%                     &  &|             c &∈ ⋃ { { c1 } if T ∈ 𝒜 (σ,a), { c2 } if F ∈ 𝒜 (σ,a) }   \\
%                     &  &}                                                                       \\
% 𝒞 <a₁(a₂,a₃),σ>     &≔ &{ <c,σ'>                                                                \\
%                     &  &| <λ(x)(k) → c> &∈ 𝒜 (σ,a₁)                                             \\
%                     &  &|            v₂ &∈ 𝒜 (σ,a₂)                                             \\
%                     &  &|            v₃ &∈ 𝒜 (σ,a₃)                                             \\
%                     &  &|            σ' &≔ σ ⊔ [x ↦ v₂] ⊔ [k ↦ v₃]                              \\
%                     &  &}                                                                       \\
% \end{align*}

The complete analysis of a program c is defined as the least fixed point of a
\emph{collection semantics} for the relation $𝒞$:
\begin{align*}
μ(Σ) → {<c,⊥>} ⊔ 𝒞*(Σ)
\end{align*}
where
\begin{align*}
𝒞*    &: \PM(Call × \AStore) → \PM(Call × \AStore) \\
𝒞*(Σ) &≔ ⋃ { 𝒞 <c,σ> | <c,σ> ∈ Σ }               \\
\end{align*}
(The collecting semantics tracks all states that the program could be in rather than just the final states.)

The first insight in monadic static analysis is to abbreviate the definitions of $𝒜 $ and $𝒞$ using a monad.
At this point, the monad "trick" is nothing more than simplifying the definition. 
This simplification is similar to how a functional programmer would use a state monad in place of explicit state passing.

We use a powerset monad and state monad transformer to write the analysis in monadic style.
The full definitions of the monads used throughout this paper are defered to section \ref{section:Proofs+Definitions}.
Because monad transformers are just fancy names for simple types, we write the simple types underneath definitions which use monad transformers.
The monadic conversion of the above analysis is:
\begin{align*}
ℳ     &: Set → Set                  \\
ℳ (a) &≔ \ST(\AStore)(\PM)(a)       \\
ℳ (a) &≔ \AStore → \PM(a × \AStore) \\
\end{align*}

>>>>

We use traditional notation for function definitions and applications rather than lambda calculus notation.
For example, we will write:
\begin{equation*}
compose(g, f, x) ≔ g(f(x))
\end{equation*}
rather than:
\begin{equation*}
compose\; g\; f\; x ≔ g\; (f\; x)
\end{equation*}

We use lambda notation for anonymous functions.
For example, the anonymous composition function is written:
\begin{equation*}
λ(g, f, x) → g(f(x))
\end{equation*}

@@

A basic familiarity with the concept of monads will be necessary for understanding this paper.
Some find it useful to review the concept of functor before learning about monads. 
We adopt this approach for our brief explanation of monads.

A type $(𝒯 : Set → Set)$ is called a \emph{functor} if one can:
\begin{itemize}
\item Define $map : ∀ (A, B : Set), (A → B) → (𝒯(A) → 𝒯(B))$
\item Prove $map(id) = id$, where $id$ is the identity function $(λ(x) → x)$
\item Prove $map (g ∘ f) = map(g) ∘ map(f)$
\end{itemize}

\paragraph{Example:} 
Lists are functors, where $map$ is defined:
\begin{align*}
    map(f)(xs) ≔ &\case(xs):           \\
          Nil    &→ Nil                \\
         x ∷ xs' &→ f(x) ∷ map(f)(xs') \\
\end{align*}

@@

Likewise a type $(𝒯 : Set → Set)$ is called a \emph{monad} if one can:
\begin{itemize}
\item Define $extend : ∀ (A, B : Set), (A → 𝒯(B)) → (𝒯(A) → 𝒯(B))$
\item Prove unit and associativity laws (not mentioned here).
\end{itemize}
The only difference from the definition of functor is in the the first argument.
For monads, this argument is allowed to be a \emph{monadic} $(A → 𝒯(B))$, rather than a pure function $(A → B)$.

\paragraph{Example:} Lists are monads, where $extend$ is defined:
\begin{align*}
    extend(f)(xs) ≔ &\case(xs):              \\
                Nil &→ Nil                   \\
            x ∷ xs' &→ f(x) ⧺ extend(f)(xs') \\
\end{align*}

@@

$extend$ for lists can also be understood as $concat$ followed by $map$:
\paragraph{Example:} 
\begin{align*}
extend(f)([10, 100]) &= concat(map(f)([10, 100]))    \\
                     &= concat([[9, 11], [99, 101]]) \\
                     &= [9, 11, 99, 101]             \\
\end{align*}
For all monads, it is equivalent to define $return$, $map$ and $join$, where:
\begin{equation*}
join : ∀ (A : Set), 𝒯(𝒯(A)) → 𝒯(A)
\end{equation*}
As we have just seen, $join$ for the list monad is just $concat$.

@@

Monad transformers are functions between monads.
Where a monad $𝒯$ will have type $Set → Set$, a monad transformer will have type $(Set → Set) → (Set → Set)$.

Monad transformers are used to extend an existing monad to support another effect.
The three monads used in this work are the state monad, powerset monad, and the identity monad.
The state monad is notated $\SM(𝓈)$ (carrying a single cell of type $𝓈$), the powerset monad is notated $\PM$, and the identity is notated $ID$.
The state and powerset monad transformer equivalents are notated $\ST(𝓈)(𝓂 )$ and $\PT(𝓂 )$ where $𝓂 $ is the underlying monad.
The state monad has $get$ and $put$ effects, whereas the powerset monad has a $nondeterminism$ effect.
The transformer versions of both of these monads allow you to combine effects piecewise.

\paragraph{Example:}
$\ST(ℤ)(\PT(ID))$ is a monad which has $get$ and $put$ effects for a single cell of integer state, in addition to nondeterminism effects.

@@

This work follows the small-step operational approach to language semantics pioneered by Plotkin and Felleisen.
In small-step operational semantics, the meaning of an expression $e$ of a language is given by the transitive closure of some step relation $e ↦ e'$.
This yields a state-machine approach to semantics.
Exploring the meaning of $e$ reduces to exploring all possible transitions reachable from $e$ through $↦$.
This contrasts to previously developed deontational approaches, where the meaning of $e$ must be given all at once by a denotation function $⟦ e ⟧$.
This denotation function can quickly demand lots of mathematical machinery.
Small-step operational methods side-step the need for such machinery.

@@

Abstract Iterpretation (AI) is a formal framework for program analysis pioneered by Cousot and Cousot.
In the setting of abstract interpretation, a program analysis is just an alternate semantics over an abstract domain.
To distinguish the original semantics from the alternate semantics we call the original semantics the \emph{concrete} semantics.
The alternate semantics is called the \emph{abstract} semantics\footnote{
  This terminology quickly becomes confusing because we will eventually construct an abstract machine for both semantics.
  This yields a concrete abstract machine and an abstract abstract machine. 
  This linguistic pun is exploited mercilessly in the work of Might and Van Horn as they “abstract” abstract abstract machines, yielding abstract abstract abstract machines.
}.

Consider a concrete language $\C$ and a program in that language $(\Ce : \C)$.
In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A relationship between $\C$ and $\A$ that defines when some $(\Ce' : \C)$ and $(\Ae' : \A)$ are related.
  This often takes the form of projecting $\A$ to $\PM(\C)$ and using the subset relation.
\item 
  An abstract version of our concrete program $\Ce$: $(\Ae : \A)$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A method to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
By convention, we put hats ($\widehat{\;\;}$) on things to name their abstract cousins.
The overall approach is summarized in the following picture:
\input{diagrams/AI.tikz}
This picture takes place at some point in the analysis process.
$\Ce$ is the current program (possibly the result of running the base program for a little while).
$\Ae$ is a valid abstraction of $\Ce$, where this validity is expressed some relation $R$ holding between $\Ce$ and $\Ae$.
$\Ce'$ is some next state of execution of $\Ce$, and likewise for $\Ae$/$\Ae'$.
In order to be a correct analysis, it must be guaranteed that $\Ce'$ and $\Ae'$ will be related.
The logical structure of the picture, which states the correctness of the analysis, is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce R \Ae) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' R \Ae')
\end{equation*}

@@

The AI setting can be elegently simplified and enriched through the use of \emph{galois connections}.
Galois connections serve as a unifying framework for establishing the “relationship between $\C $ and $\A$” mentioned in the previous section.

A galois connection between two posets (sets with a partial order) $\C$ and $\A$ is notated $\C\galois{α}{γ}\A$ and contains:
\begin{itemize}
\item $(α : \C → \A)$ where $α$ is monotonic
\item $(γ : \A → \C )$ where $γ$ is monotonic
\item A proof that $(γ ∘ α)$ is expansive: $∀ (x : \C ), x ⊑ γ(α(x))$
\item A proof that $(α ∘ γ)$ is contractive: $∀ (y : \A), α(γ(y)) ⊑ y$
\end{itemize}
The last two properties can be succinctly stated as $(α ∘ γ ⊑ id ⊑ γ ∘ α)$\footnote{
  This uses the logical monotonicity relation $f ⊑ g ⇔  (x ⊑ y ⇒  f(x) ⊑ g(y))$ for the function space.
}.
Equivalent to all four properties is the property $x ⊑ γ(y) ⇔  α(x) ⊑ y$.

\paragraph{Example:}
Given a galois connection $\C\galois{α}{γ}\A$, there exists a galois connection $(\C → \C)\galois{α'}{γ'}(\A → \A)$ where:
\begin{align*}
α'(f : \C → \C) &≔ α ∘ f ∘ γ \\
γ'(g : \A → \A) &≔ γ ∘ g ∘ α \\
\end{align*}

\paragraph{Example:} 
The language $(\PM(ℤ),+,*)$ forms a galois connection with the language $(\PM(\{ EVEN, ODD \}),∧,∨)$ where:
\begin{align*}
α(zs : \PM(ℤ))               &≔ ⋃ \{ \{ EVEN  \;|\; ∃ z ∈ zs ∧ Even(z) \},  \{ ODD   \;|\; ∃ z ∈ zs ∧ Odd(z) \} \} \\
γ(ts : \PM(\{ EVEN, ODD \})) &≔ ⋃ \{ \{ z ∈ ℤ \;|\; EVEN ∈ ts ∧ Even(z) \}, \{ z ∈ ℤ \;|\; ODD ∈ ts ∧ Odd(z) \} \} \\
\end{align*}

Galois connections simplify the AI framework by using $x ⊑ γ(y)$ or (equivalently) $α(x) ⊑ y$ as the relation $(x R y)$.
Galois connections are a natural and general way of placing partial orders \emph{on sets themselves}.
$x ⊑ γ(y)$ can be seen as a heterogenous extension of $x ⊑ y$ when $(x : A)$ and $(y : B)$ live in different sets. 
This heterogenous order is given meaning when there exists a galois connection $A\galois{α}{γ}B$.
One can also think of galois connections as something like an isomorphism, but with a weaker round-trip property.
(An isomorphism would require $α ∘ γ = id = γ ∘ α$.)

Using galois connections, the AI framework introduced in the previous section can be re-stated. In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A galois connection $\C\galois{α}{γ}\A$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A way to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
The overall approach is summarized in the following picture:
\input{diagrams/AIGC.tikz}
As before, the logical structure of the picture is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce ⊑ γ(\Ae)) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' ⊑ γ(\Ae'))
\end{equation*}
The statement of this property can be simplified further as $\Csteps ⊑ γ(\Asteps)$.
This uses the definition of galois connections for function spaces shown in a previous example.

Using a galois connection $\C\galois{α}{γ}\A$ and abstract step function $\Asteps ⊑ γ(\Csteps)$, the analysis process becomes more specific:
\begin{itemize}
\item Translate $(\Ce : \C)$ to $(α(\Ce) : \A)$
\item Explore all abstract states $\Ae'$ states reachable from $α(\Ce)$.
\item All reachable concrete states are summarized by projecting each $γ(\Ae')$ back to $\C$.
\end{itemize}

The galois connection framework simultaneously guarantees the \emph{soundness} and \emph{tightness} of this method.
Soundness tells us that if we abstract, take an abstract step, and then concretize, that the result is an approximation of taking a concrete step.
Soundness tells us that we can trust the results of the analysis.
\input{diagrams/GCSoundness.tikz}
Tightness tells us that if we concretize, take a concrete step, and then abstract, then the result must be just as precise as taking an abstract step.
Tightness tells us that our abstract step function isn't losing precision unnecessarily, and that our abstract step $\Asteps$ is provably as strong as possible.
\input{diagrams/GCTightness.tikz}

@@

We define the CPS-IF language as follows:
\input{snips/CPS-IF.tex}

@@

Control flow analysis is a class of analysis which is particularly important for higher-order languages.
In non-higher-order languages, it is useful to distinguish control-flow from data-flow.
The control flow of the program is a graph of what functions are called from where.
The data flow of a program is a mapping of which values can flow to which variables.
Traditionally one performs a control-flow analysis first, to find out which functions are called, and a data-flow analysis second, using the results of the control flow analysis.

In higher-order languages, data-flow and control-flow are tightly coupled.
Before you can tell which functions will be called, you need to know how values flow, because functions are themselves values.

We note that the distinction between “higher-order” languages and “non-higher-order” languages is a red herring.
Functions can be passed as values in C using function pointers, and object-oriented languages enjoy a similar circularity between control- and data- flow due to method dispatch.
You can read these statements as saying “all languages are higher order, and thus all languages need higher order control flow analysis”.
Or you can read them as saying “all languages are higher order, and we've done just fine without higher oder control flow analysis so far”.
The real distinction is in the code that you end up analysing.

We build on the tradition of control flow analysis (CFA) pioneered by Shivers, as well as recent refinements of CFA developed by Might and Van Horn.
In this tradition, the result of a control flow analysis is closer to what one would think of as a data-flow analysis, only the process accounts for higher order flow.
0CFA, the most basic control flow analysis, merely computes the set of values which might flow to a particular variable.
kCFA is a suite of context sensitive extensions to 0CFA for which functions are analyzed separately for each possible calling context.
We use 0CFA as the example analysis in this paper.
However our framework, and our implementation of it, scale to the full range of context sensitive control flow analyses.

@@

Monadic abstract interpretation[Sergey et al. PLDI 2013] demonstrated that abstract interpreters can be built modularly using a monadic abstraction.
We build directly on this work at the level of intuition—monads serve as our pivot-point for achieving modularity.
However, in our pursuit of a modular correctness framework, our approach deviates greatly from prior work in the specific interfaces used.

First we demonstrate a simple 0CFA analysis on CPS-IF which doesn't use our framework.
The CPS-IF language is introduced in section \ref{section:Background:ContinuationPassingStyle}.
We repeat the definition here for convenience.
\input{snips/CPS-IF.tex}

Our abstract semantics of 0CFA will track literals (including lambdas) that appear in the program text.
For integers that do not appear in the program text, the analysis uses a single token $INT$ which conservatively approximates any possible integer.
Formally, the state space $Σ$ for the abstract machine of the abstract semantics is defined as:
\begin{align*}
v : \AVal   &⩴ \INT \;|\; l \;|\; \lam(x) → c \;|\; \lam(x,k) → c \\
σ : \AStore &≔ Var → \PM(\AVal)                                   \\
Σ           &≔ \PM(Call × \AStore)                                \\
\end{align*}
Because there are a finite many literals in the program text, we can claim there are a finite number of possible abstract values.

The abstract semantics for $Σ$ comes in two parts.  
First we define a \emph{denotation function} $𝒜 $ for $Atom$ expressions.
\begin{align*}
𝒜                            &: \AStore × Atom → \PM(\AVal) \\
𝒜 (σ,x)                      &≔ σ(x)                        \\
𝒜 (σ,l)                      &≔ \{ l \}                     \\
𝒜 (σ,\add a) | 𝒜 (σ,a) ⊑ INT &≔ \{ \INT \}                  \\
𝒜 (σ,\sub a) | 𝒜 (σ,a) ⊑ INT &≔ \{ \INT \}                  \\
𝒜 (σ,\gez a) | 𝒜 (σ,a) ⊑ INT &≔ \{ \TRUE , \FALSE \}        \\
𝒜 (σ,\lam(x) → c)            &≔ \{ \lam(x) → c \}           \\
𝒜 (σ,\lam(x)(k) → c)         &≔ \{ \lam(x)(k) → c \}        \\
\end{align*}

Second we define a \emph{step relation} (as a function) $𝒞$ for $Call$ expressions.
\begin{align*}
𝒞                        : &Call × \AStore → \PM(Call × \AStore)                                              \\
𝒞(\iif(a)\{c₁\}\{c₂\},σ) ≔ &\{ (c,σ)                                                                          \\
                           &|\; c ∈ ⋃ \{ \{ c₁ \;|\; \TRUE ∈ 𝒜 (σ,a) \}, \{ c₂ \;|\; \FALSE ∈ 𝒜 (σ,a) \} \} \\
                           &\}                                                                                \\
𝒞(a₁(a₂,a₃),σ)           ≔ &\{ (c,σ')                                                                         \\
                           &|\; (\lam(x)(k) → c) ∈ 𝒜 (σ,a₁)                                                   \\
                           &|\;             v₂ ∈ 𝒜 (σ,a₂)                                                     \\
                           &|\;             v₃ ∈ 𝒜 (σ,a₃)                                                     \\
                           &|\;             σ' ≔ σ ⊔ [x ↦ v₂] ⊔ [k ↦ v₃]                                      \\
                           &\}                                                                                \\
𝒞(\halt(a),σ)            ≔ &\{ (\halt(a),σ) \}                                                                \\
\end{align*}
(We omit the $(a₁(a₂))$ case. It is directly analogous to $(a₁(a₂,a₃))$).

The complete analysis of a program $c$ is defined as the least fixed point of a \emph{collection semantics} for the relation $𝒞$:
\begin{align*}
\text{analysis} ≔ μ(Σ) → \{(c,⊥)\} ⊔ 𝒞^{⋆}(Σ)
\end{align*}
where
\begin{align*}
𝒞^{⋆}    &: \PM(Call × \AStore) → \PM(Call × \AStore) \\
𝒞^{⋆}(Σ) &≔ ⋃ \{ 𝒞(c,σ) | (c,σ) ∈ Σ \}                  \\
\end{align*}
The collecting semantics tracks all states that the program could be in rather than just the final states.

The first insight in monadic static analysis is to abbreviate the definitions of $𝒜 $ and $𝒞$ using a monad.
At this point, the monad “trick” is nothing more than a technique to simplify the definition of $𝒜 $ and $𝒞$.
This simplification is similar to how a functional programmer would use a state monad in place of state passing style.

We wll use a powerset monad and state monad transformer to write the analysis in monadic style.
The full definitions of the monads (and their proofs) used throughout this paper are defered to section \ref{section:Proofs+Definitions}.
Because monad transformers are just fancy names for simple types, we write the equivalent simple type underneath definitions which reference monad transformer.
The monadic conversion of the above analysis is as follows.
\begin{align*}
ℳ     &: Set → Set                  \\
ℳ (a) &≔ \ST(\AStore)(\PM)(a)       \\
ℳ (a) &≔ \AStore → \PM(a × \AStore) \\
\end{align*}
